---
title: "Drug overdose and alcohol related deaths"
author: 
- Dennis Campoverde-Lema
- Samantha Parras 
- Jodie Chen 
date: "July 26, 2024"
toc: true
format:
  html:
    theme: cosmo
    html-math-method: katex
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

# Introduction

  In 2022, 46.8 million Americans who were twelve years of age or older struggled with a drug use disorder.  Eight million Americans aged twelve and older battled alcohol and drug use issues at the same time that year. 

  Use of drugs and alcohol has substantial financial consequences. Every year, excessive alcohol use costs 249 billion dollars in lost productivity, medical expenditures, and criminal justice costs. In contrast, illicit drug usage costs 193 billion dollars in lost productivity, criminality, and medical costs. Communities can benefit from adopting and putting into practice policies that lessen prescription drug abuse and excessive alcohol consumption.


## Our Question

**Are there demographic and social factors that are predictors of drug overdose and alcohol-related incidents (e.g., driving accidents)?**

# Data

The data is collected by the University of Wisconsin Population Health Institute. It assigns a ranking based on health outcomes and health factors to each county in every state. There were originally 770 variables in the data set. Since the data set is so big, another data set was created with only the variables related to drug overdose deaths and alcohol-impaired driving deaths. Predictor variables are drug overdose deaths and alcohol-impaired driving deaths.

Here are some of the predictor variables picked and the reason for selecting them:

**Unemployment:** Compared to the employed population, the unemployed population has worse health and higher death rates. 

**Median Income:** The impact of median household income on the likelihood of poverty on individuals and families can have detrimental effects on their mental and physical well-being.

**Disconnected Youth:** Disconnected youth have a higher likelihood of experiencing violence, smoking, drinking, and using marijuana.

**High School Graduation:** A higher level of education is linked to a lower likelihood of smoking. Adults with higher levels of education typically have more employment opportunities and make more money overall.

**Social Associations:** Reduced participation in communal life and little social interaction with others are linked to higher rates of illness and early death. Compared to those with strong networks, those without strong social networks are less likely to choose healthy lifestyles.

(source: https://www.countyhealthrankings.org/health-data)


```{r} 
library(janitor)
library(ggplot2)
library(tidyverse)
library(sf)
library(dplyr)
library(broom)
library(VIM)
library(ggfortify)
library(class)


```


```{r, include=FALSE}

health_data_alc <- read_csv("data/final_data.csv")

health_data_alc <- health_data_alc |> 
  select(-name, -...1, -population, -drug_overdose_deaths,-homicides) |> 
  filter(!is.na(alcohol_impaired_driving_deaths))
 
health_data_alc <- na.omit(health_data_alc)


health_data_drug <- read_csv("data/final_data.csv")

health_data_drug <- health_data_drug |> 
  select(-name, -...1, -population, -alcohol_impaired_driving_deaths,-homicides) |> 
  filter(!is.na(drug_overdose_deaths))

health_data_drug <- na.omit(health_data_drug)

```
 
Upon inspection, the variable of drug overdose deaths was found to have many N/A values, and it is important to ensure this variable is clean as it is one of the main variables to be analyzed. Additionally, rename the data so that it can be read easier and look presentable for any visuals. 

```{r, include=FALSE} 

```

# Intial EDA 

Looking at the states with respect to drug overdose deaths and alcohol-impaired driving deaths, the areas with the most issues can be identified, and solutions can eventually be developed.

```{r, include=FALSE} 
health_data <- read.csv("data/analytic_data2024.csv") %>%  
  clean_names()

selected_data <- health_data %>% 
  select(
    state_fips_code, county_fips_code, state_abbreviation, name,
    drug_overdose_deaths_raw_value, alcohol_impaired_driving_deaths_raw_value,
    unemployment_raw_value, income_inequality_raw_value, high_school_completion_raw_value, 
    adult_smoking_raw_value, excessive_drinking_raw_value, median_household_income_raw_value,
    some_college_raw_value, children_in_poverty_raw_value, children_in_single_parent_households_raw_value,
    social_associations_raw_value, x_below_18_years_of_age_raw_value, 
    x_65_and_older_raw_value, x_non_hispanic_black_raw_value, 
    x_american_indian_or_alaska_native_raw_value, x_asian_raw_value, 
    x_native_hawaiian_or_other_pacific_islander_raw_value, x_hispanic_raw_value, 
    x_non_hispanic_white_raw_value, x_not_proficient_in_english_raw_value, 
    x_female_raw_value, x_rural_raw_value, population_raw_value
  )

selected_data <- selected_data |> 
  mutate(across(
    starts_with("drug_overdose_deaths_raw_value") | starts_with("alcohol_impaired_driving_deaths_raw_value") |
      starts_with("unemployment_raw_value") | starts_with("income_inequality_raw_value") | 
      starts_with("high_school_completion_raw_value") | starts_with("adult_smoking_raw_value") |
      starts_with("excessive_drinking_raw_value") | starts_with("median_household_income_raw_value") |
      starts_with("some_college_raw_value") | starts_with("children_in_poverty_raw_value") | 
      starts_with("children_in_single_parent_households_raw_value") | 
      starts_with("social_associations_raw_value") | starts_with("x_below_18_years_of_age_raw_value") | 
      starts_with("x_65_and_older_raw_value") | starts_with("x_non_hispanic_black_raw_value") | 
      starts_with("x_american_indian_or_alaska_native_raw_value") | starts_with("x_asian_raw_value") | 
      starts_with("x_native_hawaiian_or_other_pacific_islander_raw_value") | starts_with("x_hispanic_raw_value") | 
      starts_with("x_non_hispanic_white_raw_value") | starts_with("x_not_proficient_in_english_raw_value") | 
      starts_with("x_female_raw_value") | starts_with("x_rural_raw_value") | starts_with("population_raw_value"),
    as.numeric
  ))
```


```{r} 
library(maps)
state_data <- selected_data |>
  filter(county_fips_code=='000',state_abbreviation!="US", state_abbreviation !="DC") |>  
  rename(state=name) 
state_data$state <- tolower(state_data$state)

us_states_map <- map_data("state")

merged_data <- us_states_map %>%
  left_join(state_data, by = c("region" = "state"))

ggplot(data = merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = drug_overdose_deaths_raw_value), color = "white") +
  scale_fill_gradient(low = "blue", high = "red", na.value = "grey50") +
  coord_fixed(1.3)+
 theme_minimal() +  # Apply the minimal theme
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(), 
    plot.title = element_text(hjust = 0.5, face = "bold", size = 20),  
    plot.subtitle = element_text(hjust = 0.5, face = "plain", size = 8),   
    plot.caption = element_text(face = "italic", size = 6, hjust = 0)  
  ) +
  labs(
    title = "Drug Overdose Deaths by State",
    subtitle = "Data source: County Health Rankings & Roadmaps (https://www.countyhealthrankings.org/)",
    fill = "*Drug Overdose Deaths"
  )


ggplot(data = merged_data, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = alcohol_impaired_driving_deaths_raw_value), color = "white") +
  scale_fill_gradient(low = "blue", high = "red", na.value = "grey50",labels = scales::percent_format(scale = 100)) +
  coord_fixed(1.3) +
  theme_minimal() +  # Apply the minimal theme
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),  
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),  
    plot.subtitle = element_text(hjust = 0.5, face = "plain", size = 8),  
    plot.caption = element_text(face = "italic", size = 6, hjust = 0))+
  labs(
    title = "Alcohol Impaired Driving Deaths by State",
    subtitle = "Data source: County Health Rankings & Roadmaps (https://www.countyhealthrankings.org/)",
    fill = "*Alcohol Driving Deaths",
    caption = "*Percentage of motor vehicle crash deaths with alcohol involvement."
  )


```

West Virginia has the highest amount of drug overdose deaths. By conducting further research on West Virginia, it was found that the state leads the nation in opioid-related drug overdose deaths. According to the Drug Enforcement Administration, West Virginia has one of the highest prescription rates for opioids in the country and also has a high prevalence of controlled prescription drug addiction and trafficking. Furthermore, the usage of illegal prescription drugs was a factor in almost 61% of state overdose deaths in 2015.

For alcohol-impaired driving deaths, Montana had the highest amount of deaths resulting from alcohol-impaired driving. According to Forbes, Montana has the highest rate of drunk driving, with 8.57 fatal crashes involving drunk drivers for every 100,000 licensed drivers and 7.14 fatal crashes involving drunk drivers for every 100,000 state population. These rates are the highest in the country. Montana has the highest percentage of drunk drivers in the country—more than two-fifths (43.51%) of road fatalities.

By analyzing the predictors of both drug overdose deaths and alcohol-impaired driving deaths, solutions and regulations can hopefully be developed to help these states with their ongoing problems. 


# Modeling

To eventually uncover the predictors, the first step is to conduct a regression analysis. This can assist in deciding which variables should be taken into consideration and which ones not. First, finding the best regression model for the data is essential, as this can help achieve more accurate results.


**Test and Train to find the best model for our data** 

First looking at drug overdose: 

```{r}

library(dplyr)
library(glmnet)
library(tidyr)


set.seed(101)
k <- 5
health_data_drug <- health_data_drug |>
  mutate(test_fold = sample(rep(1:k, length.out = n())))

get_test_pred <- function(k) {
  test_data <- health_data_drug |> filter(test_fold == k)                     # get test and training data
  train_data <- health_data_drug |> filter(test_fold != k)
  test_x <- as.matrix(select(test_data, -drug_overdose_deaths))                       # get test and training matrices
  train_x <- as.matrix(select(train_data, -drug_overdose_deaths))
  
  lm_fit <- lm(drug_overdose_deaths ~ ., data = train_data)                           # fit models to training data
  ridge_fit <- cv.glmnet(train_x, train_data$drug_overdose_deaths, alpha = 0)
  lasso_fit <- cv.glmnet(train_x, train_data$drug_overdose_deaths, alpha = 1)
  enet_fit <- cv.glmnet(train_x, train_data$drug_overdose_deaths, alpha = 0.5)
  
  tibble(lm_pred = predict(lm_fit, newdata = test_data),              # return test results
         ridge_pred = as.numeric(predict(ridge_fit, newx = test_x)),
         lasso_pred = as.numeric(predict(lasso_fit, newx = test_x)),
         enet_pred = as.numeric(predict(enet_fit, newx = test_x)),
         test_actual = test_data$drug_overdose_deaths,
         test_fold = k)
}

# Initialize an empty tibble to store the results
test_pred_all <- tibble()

# Loop over the values from 1 to k and apply the get_test_pred function
for (i in 1:k) {
  test_pred_all <- bind_rows(test_pred_all, get_test_pred(i))
}


test_pred_all |>
  pivot_longer(lm_pred:enet_pred, 
               names_to = "type", 
               values_to = "test_pred") |>
  group_by(type, test_fold) |>
  summarize(
    rmse = sqrt(mean((test_actual - test_pred)^2))
  ) |> 
  ggplot(aes(x = type, y = rmse)) + 
  geom_point(size = 4) +
  theme_bw() +
  stat_summary(fun = mean, geom = "point", 
               color = "red", size = 4) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               color = "red", width = 0.2) + 
  labs(title="Test and Train: Drug Overdose Deaths", 
       y = 'Root Mean Squared Error') +
  theme(
    axis.title = element_text(size = 14),
    axis.text.y = element_text(size = 14, ),
    axis.text.x = element_text(size= 14, ),
    plot.title = element_text(size = 20)
  ) +
  scale_x_discrete(labels=c("enet_pred"="Elastic","lasso_pred"="Lasso","lm_pred"="Linear","ridge_pred"="Ridge"))
```
```{r}

set.seed(101)
k <- 5
health_data_alc <- health_data_alc |>
  mutate(test_fold = sample(rep(1:k, length.out = n())))


get_test_pred <- function(k) {
  test_data <- health_data_alc |> filter(test_fold == k)                     
  train_data <- health_data_alc |> filter(test_fold != k)
  test_x <- as.matrix(select(test_data, -alcohol_impaired_driving_deaths))                     
  train_x <- as.matrix(select(train_data, -alcohol_impaired_driving_deaths))
  
  lm_fit <- lm(alcohol_impaired_driving_deaths ~ ., data = train_data)                           # fit models to training data
  ridge_fit <- cv.glmnet(train_x, train_data$alcohol_impaired_driving_deaths, alpha = 0)
  lasso_fit <- cv.glmnet(train_x, train_data$alcohol_impaired_driving_deaths, alpha = 1)
  enet_fit <- cv.glmnet(train_x, train_data$alcohol_impaired_driving_deaths, alpha = 0.5)
  
  tibble(lm_pred = predict(lm_fit, newdata = test_data),              # return test results
         ridge_pred = as.numeric(predict(ridge_fit, newx = test_x)),
         lasso_pred = as.numeric(predict(lasso_fit, newx = test_x)),
         enet_pred = as.numeric(predict(enet_fit, newx = test_x)),
         test_actual = test_data$alcohol_impaired_driving_deaths,
         test_fold = k)
}

# Initialize an empty tibble to store the results
test_pred_all <- tibble()

# Loop over the values from 1 to k and apply the get_test_pred function
for (i in 1:k) {
  test_pred_all <- bind_rows(test_pred_all, get_test_pred(i))
}

test_pred_all |>
  pivot_longer(lm_pred:enet_pred, 
               names_to = "type", 
               values_to = "test_pred") |>
  group_by(type, test_fold) |>
  summarize(
    rmse = sqrt(mean((test_actual - test_pred)^2))
  ) |> 
  ggplot(aes(x = type, y = rmse)) + 
  geom_point(size = 4) +
  stat_summary(fun = mean, geom = "point", 
               color = "red", size = 4) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               color = "red", width = 0.2) + 
  theme_bw() +
  labs( title="Test and Train: Alcohol Impaired Driving Deaths", 
         y = 'Root Mean Squared Error') +
  theme(
    axis.title = element_text(size = 14),
    axis.text.y = element_text(size = 14, ),
    axis.text.x = element_text(size= 14, ),
    plot.title = element_text(size = 20)
  ) +
  scale_x_discrete(labels=c("enet_pred"="Elastic","lasso_pred"="Lasso","lm_pred"="Linear","ridge_pred"="Ridge"))

```
**All of the models intersect, so there isn't a "best" model to work with. Therefore, a linear model will be used, as it is already being used for alcohol.**

```{r}

```

# Results

Looking at Alcohol Impaired Driving Deaths: 

```{r}

alcohol_lm <- lm(alcohol_impaired_driving_deaths ~ . , data=health_data_alc)

alcohol_lm |>
  tidy() |>
  mutate(
    term = gsub("_", " ", term),
    term = sapply(term, function(x) { paste(toupper(substring(x, 1, 1)), substring(x, 2), sep = "", collapse = " ") }), 
    term = case_when(
      p.value <= 0.001 ~ paste0(term, "***"),
      p.value <= 0.01 ~ paste0(term, "**"),
      p.value <= 0.05 ~ paste0(term, "*"),
      p.value <= 0.1 ~ paste0(term, "."),
      TRUE ~ term
    ),
    term = fct_reorder(term, estimate)
  ) |> 
  ggplot(aes(x = estimate, y = term, fill = estimate)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error),
                 height = 0.2, linewidth = 2, color = "red", show.legend = FALSE) +
  geom_point(aes(color = estimate), size = 1.8, shape = 21, fill = "black", show.legend = FALSE) +
  labs(y = NULL,x="Coefficient") +
  theme_bw() +
  labs( title="Linear Model for Alcohol Impaired Driving Deaths") +
  theme(
    axis.title = element_text(size = 15),
    axis.text.y = element_text(size = 14, ),
    axis.text = element_text(size = 15, ),
    axis.title.x = element_text(size= 11, ),
    plot.title = element_text(size = 12)
  )

```
Unemployment is the top predictor for alcohol impaired driving deaths. Then if everything is constant, when unemployment increases by 1, then alcohol impaired driving deaths will increase by the coefficient. Other significant predictors are Excessive Drinking, Suicides, and Population Density.  


Now onto drug overdose deaths: 

```{r}

drug_lm <- lm(drug_overdose_deaths ~., data=health_data_drug)

drug_lm |> 
  tidy() |>
  mutate(
    term = gsub("_", " ", term),
    term = sapply(term, function(x) { paste(toupper(substring(x, 1, 1)), substring(x, 2), sep = "", collapse = " ") }), 
    term = case_when(
      p.value <= 0.001 ~ paste0(term, "***"),
      p.value <= 0.01 ~ paste0(term, "**"),
      p.value <= 0.05 ~ paste0(term, "*"),
      p.value <= 0.1 ~ paste0(term, "."),
      TRUE ~ term
    ),
    term = fct_reorder(term, estimate)
  ) |> 
  ggplot(aes(x = estimate, y = term, fill = estimate)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error),
                 height = 0.2, linewidth = 2, color = "red", show.legend = FALSE) +
  geom_point(aes(color = estimate), size = 1.8, shape = 21, fill = "black", show.legend = FALSE) +
  labs(y = NULL, x="Coefficient") +
  theme_bw() +
  labs(title="Linear Model for Drug Overdose Deaths")+ 
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 15, ),
    plot.title = element_text(size = 13)
  )

```

The top predictor is Female. The other significant factors are Unemployment, Adult Smoking, Childern in a Single Parent Household, Median Household Income, Non Hispanic Black, American Indian or Alaska Native, and Below 18 Years of Age. 


```{r}
selected_data$`Income level` <- cut(selected_data$median_household_income_raw_value,
                                    breaks = c(-Inf, 52610, 60986, 70897, Inf),
                                    labels = c("Low", "Lower-Middle", "Upper-Middle", "High"),
                                    right = TRUE, na.rm=TRUE)

ggplot(selected_data, aes(x = adult_smoking_raw_value, y = drug_overdose_deaths_raw_value)) +
  geom_point(size = 2, alpha = 0.7, aes(color = `Income level`)) +
  geom_smooth(method = 'lm', se = TRUE, color="black") +
  theme_minimal() +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Smoking Rate Relationship With Drug Overdose",
       x = 'Adult Smoking Rate',
       y = 'Drug Overdose Deaths per 100,000 population') +
  scale_color_manual(values = c("Low" = "darkred", "Lower-Middle" = "red", "Upper-Middle" = "blue", "High" = "darkblue")) + 
  theme(
    plot.title = element_text(size = 16))



```
```{r}



```

# Discussion

  From the models, it was found that unemployment is a top predictor for both drug overdose deaths and alcohol-impaired driving deaths. So if unemployment increases, alcohol and drug consumption would most likely increase as well. 
  
> A risk factor for substance abuse and drug use disorders is unemployment. According to research, losing one's job and money can cause psychological stress. This  can then deteriorate one's health and cause temporary coping mechanisms like substance abuse. Specifically, research indicates that losing one's job is linked to an increased risk of alcohol-related health issues and the use of illegal drugs. 
> Research indicates that substance abuse problems are associated with an increased chance of unemployment because those who struggle with their use are more likely to take sick days and work less productively, both of which can lead to job loss. Studies reveal that addiction also increases the difficulty of getting hired again or finding a new job.

## Ways to Reduce Unemployment

**Career pathways programs:** Programs for career pathways provide low-skilled people with occupation-specific training for higher-skilled roles in rapidly expanding industries including information technology, advanced manufacturing, and health care. These programs integrate social services with academic and technical education; many also include work experience. 





# Conclusion 

 


