---
title: "eda"
author: "Jodie"
date: "2024-06-21"
output: html_document
---

```{r}
colnames(selected_data)
```

## R Markdown

```{r}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(knitr)
library(broom)


```

```{r}
health_data <- read.csv("data/analytic_data2024.csv") |> 
  slice(-1) |>
  clean_names()


selected_data <- health_data |> 
  select(
    state_fips_code, county_fips_code, state_abbreviation, name,
    drug_overdose_deaths_raw_value, alcohol_impaired_driving_deaths_raw_value,
    unemployment_raw_value, income_inequality_raw_value, high_school_completion_raw_value, 
    adult_smoking_raw_value, excessive_drinking_raw_value, median_household_income_raw_value,
    some_college_raw_value, children_in_poverty_raw_value, children_in_single_parent_households_raw_value,
    social_associations_raw_value, x_below_18_years_of_age_raw_value, 
    x_65_and_older_raw_value, x_non_hispanic_black_raw_value, 
    x_american_indian_or_alaska_native_raw_value, x_asian_raw_value, 
    x_native_hawaiian_or_other_pacific_islander_raw_value, x_hispanic_raw_value, 
    x_non_hispanic_white_raw_value, x_not_proficient_in_english_raw_value,
    x_female_raw_value, x_rural_raw_value, long_commute_driving_alone_raw_value, population_raw_value, traffic_volume_raw_value
  )

# Convert relevant columns to numeric
selected_data <- selected_data |> 
  mutate(across(
    starts_with("drug_overdose_deaths_raw_value") | starts_with("alcohol_impaired_driving_deaths_raw_value") |
    starts_with("unemployment_raw_value") | starts_with("income_inequality_raw_value") | 
    starts_with("high_school_completion_raw_value") | starts_with("adult_smoking_raw_value") |
    starts_with("excessive_drinking_raw_value") | starts_with("median_household_income_raw_value") |
    starts_with("some_college_raw_value") | starts_with("children_in_poverty_raw_value") | 
    starts_with("children_in_single_parent_households_raw_value") | 
    starts_with("social_associations_raw_value") | starts_with("x_below_18_years_of_age_raw_value") | 
    starts_with("x_65_and_older_raw_value") | starts_with("x_non_hispanic_black_raw_value") | 
    starts_with("x_american_indian_or_alaska_native_raw_value") | starts_with("x_asian_raw_value") | 
    starts_with("x_native_hawaiian_or_other_pacific_islander_raw_value") | starts_with("x_hispanic_raw_value") | 
    starts_with("x_non_hispanic_white_raw_value") | starts_with("x_not_proficient_in_english_raw_value") | 
    starts_with("x_female_raw_value") | starts_with("x_rural_raw_value") | starts_with("population_raw_value") | starts_with("long_commute_driving_alone_raw_value") | starts_with("traffic_volume_raw_value"),
    as.numeric
  ))
#glimpse(health_data)
glimpse(selected_data)
```

```{r}
cor(selected_data)
```

## exploring x_non_hispanic_black_raw_value, children_in_single_parent_households_raw_value, and unemployment_raw_value 

## analyzing by county level

```{r}

ggplot(selected_data, aes(x=population_raw_value)) +
  geom_histogram(fill = "blue", color = "black") +
  labs(title = "Histogram Example", x = "Values", y = "Frequency") +
  theme_minimal()

library(ggplot2)
ggplot(selected_data, aes(x = population_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_rug(alpha=0.4) +
  scale_x_log10() + # plots on log scale
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = "Population vs alcohol_impaired_deaths", x = "population", y = "alcohol_impaired_deaths") +
  theme_minimal()

```

```{r}
library(GGally)

renamed_data <- selected_data |>
  select(
    population = population_raw_value,
    long_commute = long_commute_driving_alone_raw_value,
    traffic_volume = traffic_volume_raw_value,
    alcohol_deaths = alcohol_impaired_driving_deaths_raw_value
  )

renamed_data |>
  ggpairs() +
  scale_x_log10() 

  

# selected_data |>
#   select(population_raw_value, long_commute_driving_alone_raw_value, traffic_volume_raw_value, alcohol_impaired_driving_deaths_raw_value) |>
#   ggpairs()
```


```{r}
ggplot(renamed_data, aes(x = population, y = long_commute)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_rug(alpha=0.4) +
  scale_x_log10() + # plots on log scale
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = "Population vs long_commute", x = "population", y = "long_commute") +
  theme_minimal()
```

## drug overdose deaths and median_house_income as confounding variable for different races

```{r}
# name_mapping <- c(
#   "drug_overdose_deaths_raw_value" = "overdose_deaths",
#   "median_household_income_raw_value" = "median_income",
#   "x_asian_raw_value" = "asian_pop",
#   "x_american_indian_or_alaska_native_raw_value" = "native_pop"
# )
# 
# names(selected_data) <- name_mapping[names(selected_data)]

# short_names <- c(
#   "overdose_deaths",
#   "median_income",
#   "asian_pop",
#   "native_pop"
# )
# 
# attach(selected_data)
# pairs(drug_overdose_deaths_raw_value~median_household_income_raw_value+x_asian_raw_value+x_american_indian_or_alaska_native_raw_value, labels=short_names)


# children in poverty--percentage of people under age 18 in poverty
# does not necessarily mean people under 18 are drug overdosing
# Percentage of adults who are current smokers (age-adjusted).

# variables to not look at
# excessive drinking is only percentage of adults that report binge or heavy drinking


renamed_data2 <- selected_data |>
  select(
    median_household_income_raw_value,
    x_asian_raw_value,
    x_hispanic_raw_value,
    drug_overdose_deaths_raw_value
  )

renamed_data2 |>
  ggpairs() +
  scale_x_log10() 

ggplot(selected_data, aes(x = median_household_income_raw_value, y = drug_overdose_deaths_raw_value)) +
  scale_x_log10() +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth() +
  geom_rug(alpha=0.4) +
  labs(x = "median house income", y = "drug overdose deaths") +
  theme_minimal()

ggplot(selected_data, aes(x = log(median_household_income_raw_value), y = drug_overdose_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_rug(alpha=0.4) +
  labs(title = "median_house income", x = "log(median house income)", y = "drug overdose") +
  theme_minimal()

# to get pairs of all variables we are interested in looking at 

set2<-selected_data |>
  select(median_household_income_raw_value,
         excessive_drinking_raw_value, some_college_raw_value,
         adult_smoking_raw_value, drug_overdose_deaths_raw_value) |>
  rename(
    overdose_deaths = drug_overdose_deaths_raw_value,
    household_income = median_household_income_raw_value,
    excessive_drinking = excessive_drinking_raw_value,
    some_college_raw_value = completed_college,
    adult_smoking = adult_smoking_raw_value
  )

set2 |>
  ggpairs() 
  
```

```{r}
# looking at states only, filter by states 
selected_data<-selected_data |>
  filter(county_fips_code=="000")

glimpse(selected_data)

health_data |>
  select(traffic_volume_raw_value)

health_data |>
  select(long_commute_driving_alone_raw_value) # ranges from 0 to 1 so percentage of people with long commute

ggplot(health_data, aes(x=population_raw_value)) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of % Below 18 Years of Age", x = "% Below 18 Years of Age", y = "Frequency")
```

## alcohol impaired death--pop over 65, alaskan native, single parent house, % of people that live in census defined rural area

```{r}
set3<-selected_data |>
  select(x_65_and_older_raw_value,
         x_rural_raw_value, x_american_indian_or_alaska_native_raw_value, excessive_drinking_raw_value,
         alcohol_impaired_driving_deaths_raw_value) |>
  rename(
    overdose_deaths = drug_overdose_deaths_raw_value,
    household_income = median_household_income_raw_value,
    excessive_drinking = excessive_drinking_raw_value,
    some_college_raw_value = completed_college,
    adult_smoking = adult_smoking_raw_value
  )

set3 |>
  ggpairs() +
  scale_x_log10()

ggplot(selected_data, aes(x=log(x_american_indian_or_alaska_native_raw_value), y=alcohol_impaired_driving_deaths_raw_value))  +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE)

library(car)

x_rural_raw_value <-x_rural_raw_value + 1
x_american_indian_or_alaska_native_raw_value<-x_american_indian_or_alaska_native_raw_value + 1

# power transform doesn't work unless u have nonzero values
summary(powerTransform(cbind(x_65_and_older_raw_value,
         x_rural_raw_value, x_american_indian_or_alaska_native_raw_value)~1))
```

```{r}
ggplot(selected_data, aes(x = x_female_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = 'Female Population vs Alcohol-Impaired Driving Deaths', x = 'Female Population', y = 'Alcohol-Impaired Driving Deaths')

lin_model<-lm(alcohol_impaired_driving_deaths_raw_value~x_female_raw_value, data=selected_data)

summary(lin_model)

ggplot(selected_data, aes(x = log(x_american_indian_or_alaska_native_raw_value), y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'american indian or alaska natives', y = 'Alcohol-Impaired Driving Deaths')

ggplot(selected_data, aes(x = x_american_indian_or_alaska_native_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  scale_x_log10() +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'american indian or alaska natives', y = 'Alcohol-Impaired Driving Deaths')

```

## analyzing excessive drinking vs alcohol-impaired driving deaths

```{r}
ggplot(selected_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(color="darkred", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  scale_x_log10() +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'excessive drinking rate', y = 'Alcohol-Impaired Driving Deaths')
```

```{r}
# below age of 18: raw percentage of people below 18 years old in the state 
# 65 and older raw value: raw percentage of peple 65 and older
# alcohol-impaired driving deaths
# excessive drinking raw value (percenatage)

# see which state has the highest alcohol-impaired deaths

ggplot(selected_data, aes(x=state_abbreviation, y=alcohol_impaired_driving_deaths_raw_value)) +
  geom_bar(stat="identity")
```

```{r}
# issue with git?
# clustering analysis 
# alcohol impaired driving deaths
# x_female_raw_value is % female raw value 
# Q: what does the numerator and denominator mean vs the raw value more generally
# graphs that allow you to plot confidence intervals
# error in plotting the histogram for below

# for social factors--what amount median household income?
selected_data |>
  select(x_female_raw_value)
```

```{r}
health_data |>
  select(population_raw_value, population_numerator, population_denominator)

rawnum <- health_data |>
  select(v051_numerator)

```

```{r}
# Histogram for % Below 18 Years of Age
ggplot(selected_data, aes(x = )) +
  geom_histogram(bins = 50, fill = "blue", alpha = 0.7) +
  labs(title = "Distribution of % Below 18 Years of Age", x = "% Below 18 Years of Age", y = "Frequency")

# Histogram for % 65 and Older
ggplot(selected_data, aes(x = x_65_and_older_raw_value)) +
  geom_histogram(bins = 50, fill = "green", alpha = 0.7) +
  labs(title = "Distribution of % 65 and Older", x = "% 65 and Older", y = "Frequency")

# Boxplot for % Below 18 Years of Age
ggplot(df, aes(x = "", y = v052_rawvalue)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  labs(title = "Box Plot of % Below 18 Years of Age", y = "% Below 18 Years of Age")

# Boxplot for % 65 and Older
ggplot(df, aes(x = "", y = v053_rawvalue)) +
  geom_boxplot(fill = "green", alpha = 0.7) +
  labs(title = "Box Plot of % 65 and Older", y = "% 65 and Older")

```

## alcohol impaired deaths based on montana state only 

```{r}

states_w_highest_alcohol_death <- selected_data |>
  filter(state_abbreviation %in% c("MT", "ND", "SD", "CO", "WI"))

states_w_lowest_alcohol_death <- selected_data |>
  filter(state_abbreviation %in% c("AZ", "KS", "ID", "FL", "GA")) # removed UTAH


MT_alcohol_death <- selected_data |>
  filter(state_abbreviation=="MT")

# color the scatterplot 
ggplot(states_w_lowest_alcohol_death, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value)) +
  geom_point(color="darkorange", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  #scale_x_log10() +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'excessive_drinking vs Alcohol-Impaired Driving Deaths', x = 'excessive drinking rate', y = 'Alcohol-Impaired Driving Deaths')
```

## Comparing counties in states with high alcohol death vs with low alcohol death

```{r}
combined_data <- bind_rows(
  states_w_highest_alcohol_death %>% mutate(Counties = "in states with High Alcohol Death"),
  states_w_lowest_alcohol_death %>% mutate(Counties = "in states with Low Alcohol Death")
)

ggplot(combined_data, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'Median Household Income vs Excessive Drinking',
       x = 'Median Household Income',
       y = 'Excessive Drinking Rate') +
  scale_color_manual(values = c("in states with High Alcohol Death" = "darkred", "in states with Low Alcohol Death" = "darkgreen"))

```

```{r}
# counties_w_highest_alcohol_death <- selected_data |>
#   filter(alcohol_impaired_driving_deaths_raw_value)
# 
# states_w_lowest_alcohol_death <- selected_data |>
#   filter(state_abbreviation %in% c("AZ", "KS", "ID", "FL", "GA")) # removed UTAH

#median_value <- median(selected_data$alcohol_impaired_driving_deaths_raw_value, na.rm = TRUE)

# selected_data <- selected_data |>
#   mutate(Counties = ifelse(alcohol_impaired_driving_deaths_raw_value > median_value, 
#                                          "High Alcohol Impaired Deaths", 
#                                          "Low Alcohol Impaired Deaths"))

high_threshold <- 0.3364
low_threshold <- 0.1818

# Create a new variable based on the thresholds
selected_data_first_and_third <- selected_data %>%
  filter(alcohol_impaired_driving_deaths_raw_value > high_threshold | alcohol_impaired_driving_deaths_raw_value < low_threshold) |>
  mutate(Counties = case_when(
    alcohol_impaired_driving_deaths_raw_value > high_threshold ~ "High Alcohol Impaired Deaths Rate",
    alcohol_impaired_driving_deaths_raw_value < low_threshold ~ "Low Alcohol Impaired Deaths Rate",
    TRUE ~ "Medium Alcohol Impaired Deaths"
  ))

ggplot(selected_data_first_and_third, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'Median Household Income vs Excessive Drinking',
       x = 'Median Household Income',
       y = 'Excessive Drinking Rate') +
  scale_color_manual(values = c("High Alcohol Impaired Deaths Rate" = "darkred", "Low Alcohol Impaired Deaths Rate" = "darkgreen"))

```

visualizing distribution of alcohol_impaired_deaths
```{r}
# Step 1: Create a named vector for the summary statistics
summary_stats <- c(Min = 0.0000, `1st Qu.` = 0.1818, Median = 0.2647, Mean = 0.2733, `3rd Qu.` = 0.3364, Max = 1.0000)

# Step 2: Create the bar plot
barplot(summary_stats, 
        main = "Summary Statistics for Alcohol Impaired Driving Deaths (Raw Value)", 
        ylab = "Values", 
        xlab = "Statistics", 
        col = c("blue", "green", "orange", "red", "purple", "brown"), 
        ylim = c(0, max(summary_stats) * 1.1), # Adding some space on top for labels
        las = 2) # Making the x-axis labels perpendicular to the axis

# Step 3: Add text labels on the bars
text(x = barplot(summary_stats, plot = FALSE), 
     y = summary_stats, 
     labels = round(summary_stats, 4), 
     pos = 3, 
     cex = 0.8, 
     col = "black")

```

## visualizing adult smoking rate vs drug overdose deaths

```{r}
summary(selected_data$drug_overdose_deaths_raw_value)

high_threshold2 <- 35.75
low_threshold2 <- 17.71

selected_data_first_and_third2 <- selected_data %>%
  filter(drug_overdose_deaths_raw_value > high_threshold2 | drug_overdose_deaths_raw_value < low_threshold2) |>
  mutate(Counties = case_when(
    drug_overdose_deaths_raw_value > high_threshold2 ~ "High Drug Overdose rate",
    drug_overdose_deaths_raw_value < low_threshold2 ~ "Low Drug Overdose rate",
    TRUE ~ "Medium Alcohol Impaired Deaths"
  ))

ggplot(selected_data_first_and_third2, aes(x = children_in_poverty_raw_value, y = adult_smoking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Counties with high overdose rates have high rate of current smokers",
       x = '% of people under 18 in poverty',
       y = '% of adult current smokers') +
  scale_color_manual(values = c("High Drug Overdose rate" = "darkred", "Low Drug Overdose rate" = "darkgreen"))

```

## visualizing excessive drinking vs median household income

```{r}
# selected_data |>
#   mutate(income_level = )

# at different states, you have different interpretations of what is considered low income, lower-middle
selected_data$income_level <- cut(selected_data$median_household_income_raw_value,
                                  breaks = c(-Inf, 52610, 60986, 70897, Inf),
                                  labels = c("Low", "Lower-Middle", "Upper-Middle", "High"),
                                  right = TRUE, na.rm=TRUE)

selected_data <- selected_data %>%
  drop_na(excessive_drinking_raw_value, alcohol_impaired_driving_deaths_raw_value)

ggplot(selected_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(size = 2, alpha = 0.5, aes(color = income_level)) +
  geom_smooth(method = 'lm', se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Counties with higher median income tend to have higher proportion of adults that excessively drink",
       x = '% of adults that excessively drink',
       y = '% of driving deaths with alcohol involvement') +
  scale_color_manual(values = c("Low" = "yellow", "Lower-Middle" = "lightgreen", "Upper-Middle" = "yellowgreen", "High" = "darkgreen")) 

# counties with higher median income tend to have higher percentage of adults that binge drink leading to higher rate of driving fatalities from alcohol

lm(alcohol_impaired_driving_deaths_raw_value~excessive_drinking_raw_value, data=selected_data)

model <- lm(alcohol_impaired_driving_deaths_raw_value ~ excessive_drinking_raw_value, data = selected_data)

# Extract the coefficients
coefficients <- coef(model)
intercept <- coefficients[1]
slope <- coefficients[2]

```

```{r}
model <- lm(alcohol_impaired_driving_deaths_raw_value ~ excessive_drinking_raw_value, data = selected_data)

# Extract the coefficients
coefficients <- coef(model)
intercept <- coefficients[1]
slope <- coefficients[2]

# Create the equation label
eq <- paste0("y = ", round(slope, 2), "x + ", round(intercept, 2))



ggplot(selected_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value, color = income_level)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', se = TRUE) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
       x = '% of adults that excessively drink',
       y = '% of driving deaths with alcohol involvement') +
  scale_color_manual(values = c("Low" = "yellow", "Lower-Middle" = "lightgreen", "Upper-Middle" = "yellowgreen", "High" = "darkgreen")) +
  stat_regline_equation(label.x = 0.1, label.y = 0.9, aes(label = ..eq.label..), size = 5)
```