---
title: "eda"
author: "Jodie"
date: "2024-06-21"
output: html_document
---

```{r}
colnames(selected_data)
```

## R Markdown

```{r}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(knitr)
library(broom)


```


```{r}
health_data <- read.csv("data/analytic_data2024.csv") |> 
  slice(-1) |>
  clean_names()


selected_data <- health_data |> 
  select(
    state_fips_code, county_fips_code, county_clustered_yes_1_no_0, state_abbreviation, name, x5_digit_fips_code,
    drug_overdose_deaths_raw_value, 
    drug_overdose_deaths_numerator, 
    drug_overdose_deaths_denominator,
    alcohol_impaired_driving_deaths_raw_value, alcohol_impaired_driving_deaths_numerator, alcohol_impaired_driving_deaths_denominator,
    unemployment_raw_value, income_inequality_raw_value, high_school_completion_raw_value, 
    adult_smoking_raw_value, excessive_drinking_raw_value, median_household_income_raw_value,
    some_college_raw_value, children_in_poverty_raw_value, children_in_single_parent_households_raw_value,
    social_associations_raw_value, x_below_18_years_of_age_raw_value, 
    x_65_and_older_raw_value, x_non_hispanic_black_raw_value, 
    x_american_indian_or_alaska_native_raw_value, x_asian_raw_value, 
    x_native_hawaiian_or_other_pacific_islander_raw_value, x_hispanic_raw_value, 
    x_non_hispanic_white_raw_value, x_not_proficient_in_english_raw_value,
    x_female_raw_value, x_rural_raw_value, long_commute_driving_alone_raw_value, population_raw_value, traffic_volume_raw_value
  )

# Convert relevant columns to numeric
selected_data <- selected_data |> 
  mutate(across(
    starts_with("drug_overdose_deaths") | starts_with("alcohol_impaired_driving_deaths") |
    starts_with("unemployment_raw_value") | starts_with("income_inequality_raw_value") | 
    starts_with("high_school_completion_raw_value") | starts_with("adult_smoking_raw_value") |
    starts_with("excessive_drinking_raw_value") | starts_with("median_household_income_raw_value") |
    starts_with("some_college_raw_value") | starts_with("children_in_poverty_raw_value") | 
    starts_with("children_in_single_parent_households_raw_value") | 
    starts_with("social_associations_raw_value") | starts_with("x_below_18_years_of_age_raw_value") | 
    starts_with("x_65_and_older_raw_value") | starts_with("x_non_hispanic_black_raw_value") | 
    starts_with("x_american_indian_or_alaska_native_raw_value") | starts_with("x_asian_raw_value") | 
    starts_with("x_native_hawaiian_or_other_pacific_islander_raw_value") | starts_with("x_hispanic_raw_value") | 
    starts_with("x_non_hispanic_white_raw_value") | starts_with("x_not_proficient_in_english_raw_value") | 
    starts_with("x_female_raw_value") | starts_with("x_rural_raw_value") | starts_with("population_raw_value") | starts_with("long_commute_driving_alone_raw_value") | starts_with("traffic_volume_raw_value"), 
    as.numeric
  )) 
#glimpse(health_data)

# add population density column
# GEOID is a unique county and corresponds to x5_digit_fips_code in health_data
library(tigris)
us_counties <- counties(cb = TRUE)
census_df <- us_counties |>
  select(x5_digit_fips_code=GEOID, land_area=ALAND)

# Inner join: only rows with matching x5_digit_fips_code will be included
new_sel <- merge(selected_data, census_df, by = "x5_digit_fips_code")

new_df <- new_sel |>
  mutate(land_area_km = land_area / 1000000) |>
  mutate(pop_den = population_raw_value / land_area_km)

glimpse(new_df) # has population density and land area_km
```

```{r}
cor(selected_data)
```

## exploring x_non_hispanic_black_raw_value, children_in_single_parent_households_raw_value, and unemployment_raw_value 

## analyzing by county level

<!-- ```{r} -->

<!-- ggplot(selected_data, aes(x=population_raw_value)) + -->
<!--   geom_histogram(fill = "blue", color = "black") + -->
<!--   labs(title = "Histogram Example", x = "Values", y = "Frequency") + -->
<!--   theme_minimal() -->

<!-- library(ggplot2) -->
<!-- ggplot(selected_data, aes(x = population_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_rug(alpha=0.4) + -->
<!--   scale_x_log10() + # plots on log scale -->
<!--   scale_y_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = "Population vs alcohol_impaired_deaths", x = "population", y = "alcohol_impaired_deaths") + -->
<!--   theme_minimal() -->

<!-- ``` -->

<!-- ```{r} -->
<!-- library(GGally) -->

<!-- renamed_data <- selected_data |> -->
<!--   select( -->
<!--     population = population_raw_value, -->
<!--     long_commute = long_commute_driving_alone_raw_value, -->
<!--     traffic_volume = traffic_volume_raw_value, -->
<!--     alcohol_deaths = alcohol_impaired_driving_deaths_raw_value -->
<!--   ) -->

<!-- renamed_data |> -->
<!--   ggpairs() + -->
<!--   scale_x_log10()  -->



<!-- # selected_data |> -->
<!-- #   select(population_raw_value, long_commute_driving_alone_raw_value, traffic_volume_raw_value, alcohol_impaired_driving_deaths_raw_value) |> -->
<!-- #   ggpairs() -->
<!-- ``` -->


<!-- ```{r} -->
<!-- ggplot(renamed_data, aes(x = population, y = long_commute)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_rug(alpha=0.4) + -->
<!--   scale_x_log10() + # plots on log scale -->
<!--   scale_y_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = "Population vs long_commute", x = "population", y = "long_commute") + -->
<!--   theme_minimal() -->
<!-- ``` -->

<!-- ## drug overdose deaths and median_house_income as confounding variable for different races -->

<!-- ```{r} -->
<!-- # name_mapping <- c( -->
<!-- #   "drug_overdose_deaths_raw_value" = "overdose_deaths", -->
<!-- #   "median_household_income_raw_value" = "median_income", -->
<!-- #   "x_asian_raw_value" = "asian_pop", -->
<!-- #   "x_american_indian_or_alaska_native_raw_value" = "native_pop" -->
<!-- # ) -->
<!-- #  -->
<!-- # names(selected_data) <- name_mapping[names(selected_data)] -->

<!-- # short_names <- c( -->
<!-- #   "overdose_deaths", -->
<!-- #   "median_income", -->
<!-- #   "asian_pop", -->
<!-- #   "native_pop" -->
<!-- # ) -->
<!-- #  -->
<!-- # attach(selected_data) -->
<!-- # pairs(drug_overdose_deaths_raw_value~median_household_income_raw_value+x_asian_raw_value+x_american_indian_or_alaska_native_raw_value, labels=short_names) -->


<!-- # children in poverty--percentage of people under age 18 in poverty -->
<!-- # does not necessarily mean people under 18 are drug overdosing -->
<!-- # Percentage of adults who are current smokers (age-adjusted). -->

<!-- # variables to not look at -->
<!-- # excessive drinking is only percentage of adults that report binge or heavy drinking -->


<!-- renamed_data2 <- selected_data |> -->
<!--   select( -->
<!--     median_household_income_raw_value, -->
<!--     x_asian_raw_value, -->
<!--     x_hispanic_raw_value, -->
<!--     drug_overdose_deaths_raw_value -->
<!--   ) -->

<!-- renamed_data2 |> -->
<!--   ggpairs() + -->
<!--   scale_x_log10()  -->

<!-- ggplot(selected_data, aes(x = median_household_income_raw_value, y = drug_overdose_deaths_raw_value)) + -->
<!--   scale_x_log10() + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth() + -->
<!--   geom_rug(alpha=0.4) + -->
<!--   labs(x = "median house income", y = "drug overdose deaths") + -->
<!--   theme_minimal() -->

<!-- ggplot(selected_data, aes(x = log(median_household_income_raw_value), y = drug_overdose_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_rug(alpha=0.4) + -->
<!--   labs(title = "median_house income", x = "log(median house income)", y = "drug overdose") + -->
<!--   theme_minimal() -->

<!-- # to get pairs of all variables we are interested in looking at  -->

<!-- set2<-selected_data |> -->
<!--   select(median_household_income_raw_value, -->
<!--          excessive_drinking_raw_value, some_college_raw_value, -->
<!--          adult_smoking_raw_value, drug_overdose_deaths_raw_value) |> -->
<!--   rename( -->
<!--     overdose_deaths = drug_overdose_deaths_raw_value, -->
<!--     household_income = median_household_income_raw_value, -->
<!--     excessive_drinking = excessive_drinking_raw_value, -->
<!--     some_college_raw_value = completed_college, -->
<!--     adult_smoking = adult_smoking_raw_value -->
<!--   ) -->

<!-- set2 |> -->
<!--   ggpairs()  -->

<!-- ``` -->

<!-- ```{r} -->
<!-- # looking at states only, filter by states  -->
<!-- selected_data<-selected_data |> -->
<!--   filter(county_fips_code=="000") -->

<!-- glimpse(selected_data) -->

<!-- health_data |> -->
<!--   select(traffic_volume_raw_value) -->

<!-- health_data |> -->
<!--   select(long_commute_driving_alone_raw_value) # ranges from 0 to 1 so percentage of people with long commute -->

<!-- ggplot(health_data, aes(x=population_raw_value)) + -->
<!--   geom_histogram(bins = 50, fill = "blue", alpha = 0.7) + -->
<!--   labs(title = "Distribution of % Below 18 Years of Age", x = "% Below 18 Years of Age", y = "Frequency") -->
<!-- ``` -->

<!-- ## alcohol impaired death--pop over 65, alaskan native, single parent house, % of people that live in census defined rural area -->

<!-- ```{r} -->
<!-- set3<-selected_data |> -->
<!--   select(x_65_and_older_raw_value, -->
<!--          x_rural_raw_value, x_american_indian_or_alaska_native_raw_value, excessive_drinking_raw_value, -->
<!--          alcohol_impaired_driving_deaths_raw_value) |> -->
<!--   rename( -->
<!--     overdose_deaths = drug_overdose_deaths_raw_value, -->
<!--     household_income = median_household_income_raw_value, -->
<!--     excessive_drinking = excessive_drinking_raw_value, -->
<!--     some_college_raw_value = completed_college, -->
<!--     adult_smoking = adult_smoking_raw_value -->
<!--   ) -->

<!-- set3 |> -->
<!--   ggpairs() + -->
<!--   scale_x_log10() -->

<!-- ggplot(selected_data, aes(x=log(x_american_indian_or_alaska_native_raw_value), y=alcohol_impaired_driving_deaths_raw_value))  + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth(method = 'lm', color = 'blue', se = TRUE) -->

<!-- library(car) -->

<!-- x_rural_raw_value <-x_rural_raw_value + 1 -->
<!-- x_american_indian_or_alaska_native_raw_value<-x_american_indian_or_alaska_native_raw_value + 1 -->

<!-- # power transform doesn't work unless u have nonzero values -->
<!-- summary(powerTransform(cbind(x_65_and_older_raw_value, -->
<!--          x_rural_raw_value, x_american_indian_or_alaska_native_raw_value)~1)) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ggplot(selected_data, aes(x = x_female_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth(method = 'lm', color = 'blue', se = TRUE) + -->
<!--   scale_x_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = 'Female Population vs Alcohol-Impaired Driving Deaths', x = 'Female Population', y = 'Alcohol-Impaired Driving Deaths') -->

<!-- lin_model<-lm(alcohol_impaired_driving_deaths_raw_value~x_female_raw_value, data=selected_data) -->

<!-- summary(lin_model) -->

<!-- ggplot(selected_data, aes(x = log(x_american_indian_or_alaska_native_raw_value), y = alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth(method = 'lm', color = 'blue', se = TRUE) + -->
<!--   scale_y_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'american indian or alaska natives', y = 'Alcohol-Impaired Driving Deaths') -->

<!-- ggplot(selected_data, aes(x = x_american_indian_or_alaska_native_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth(method = 'lm', color = 'blue', se = TRUE) + -->
<!--   scale_x_log10() + -->
<!--   scale_y_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'american indian or alaska natives', y = 'Alcohol-Impaired Driving Deaths') -->

<!-- ``` -->

<!-- ## analyzing excessive drinking vs alcohol-impaired driving deaths -->

<!-- ```{r} -->
<!-- ggplot(selected_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_point(color="darkred", size=2, alpha=0.5) + -->
<!--   geom_smooth(method = 'lm', color = 'blue', se = TRUE) + -->
<!--   scale_x_log10() + -->
<!--   scale_y_continuous(labels=scales::percent_format()) + -->
<!--   labs(title = 'alaska natives vs Alcohol-Impaired Driving Deaths', x = 'excessive drinking rate', y = 'Alcohol-Impaired Driving Deaths') -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # below age of 18: raw percentage of people below 18 years old in the state  -->
<!-- # 65 and older raw value: raw percentage of peple 65 and older -->
<!-- # alcohol-impaired driving deaths -->
<!-- # excessive drinking raw value (percenatage) -->

<!-- # see which state has the highest alcohol-impaired deaths -->

<!-- ggplot(selected_data, aes(x=state_abbreviation, y=alcohol_impaired_driving_deaths_raw_value)) + -->
<!--   geom_bar(stat="identity") -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # issue with git? -->
<!-- # clustering analysis  -->
<!-- # alcohol impaired driving deaths -->
<!-- # x_female_raw_value is % female raw value  -->
<!-- # Q: what does the numerator and denominator mean vs the raw value more generally -->
<!-- # graphs that allow you to plot confidence intervals -->
<!-- # error in plotting the histogram for below -->

<!-- # for social factors--what amount median household income? -->
<!-- selected_data |> -->
<!--   select(x_female_raw_value) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- health_data |> -->
<!--   select(population_raw_value, population_numerator, population_denominator) -->

<!-- rawnum <- health_data |> -->
<!--   select(v051_numerator) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- # Histogram for % Below 18 Years of Age -->
<!-- ggplot(selected_data, aes(x = )) + -->
<!--   geom_histogram(bins = 50, fill = "blue", alpha = 0.7) + -->
<!--   labs(title = "Distribution of % Below 18 Years of Age", x = "% Below 18 Years of Age", y = "Frequency") -->

<!-- # Histogram for % 65 and Older -->
<!-- ggplot(selected_data, aes(x = x_65_and_older_raw_value)) + -->
<!--   geom_histogram(bins = 50, fill = "green", alpha = 0.7) + -->
<!--   labs(title = "Distribution of % 65 and Older", x = "% 65 and Older", y = "Frequency") -->

<!-- # Boxplot for % Below 18 Years of Age -->
<!-- ggplot(df, aes(x = "", y = v052_rawvalue)) + -->
<!--   geom_boxplot(fill = "blue", alpha = 0.7) + -->
<!--   labs(title = "Box Plot of % Below 18 Years of Age", y = "% Below 18 Years of Age") -->

<!-- # Boxplot for % 65 and Older -->
<!-- ggplot(df, aes(x = "", y = v053_rawvalue)) + -->
<!--   geom_boxplot(fill = "green", alpha = 0.7) + -->
<!--   labs(title = "Box Plot of % 65 and Older", y = "% 65 and Older") -->

<!-- ``` -->

## alcohol impaired deaths based on montana state only 

```{r}

states_w_highest_alcohol_death <- selected_data |>
  filter(state_abbreviation %in% c("MT", "ND", "SD", "CO", "WI"))

states_w_lowest_alcohol_death <- selected_data |>
  filter(state_abbreviation %in% c("AZ", "KS", "ID", "FL", "GA")) # removed UTAH


MT_alcohol_death <- selected_data |>
  filter(state_abbreviation=="MT")

# color the scatterplot 
ggplot(states_w_lowest_alcohol_death, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value)) +
  geom_point(color="darkorange", size=2, alpha=0.5) +
  geom_smooth(method = 'lm', color = 'blue', se = TRUE) +
  #scale_x_log10() +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'excessive_drinking vs Alcohol-Impaired Driving Deaths', x = 'excessive drinking rate', y = 'Alcohol-Impaired Driving Deaths')
```

## Comparing counties in states with high alcohol death vs with low alcohol death

```{r}
combined_data <- bind_rows(
  states_w_highest_alcohol_death %>% mutate(Counties = "in states with High Alcohol Death"),
  states_w_lowest_alcohol_death %>% mutate(Counties = "in states with Low Alcohol Death")
)

ggplot(combined_data, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'Median Household Income vs Excessive Drinking',
       x = 'Median Household Income',
       y = 'Excessive Drinking Rate') +
  scale_color_manual(values = c("in states with High Alcohol Death" = "darkred", "in states with Low Alcohol Death" = "darkgreen"))

```

```{r}
# counties_w_highest_alcohol_death <- selected_data |>
#   filter(alcohol_impaired_driving_deaths_raw_value)
# 
# states_w_lowest_alcohol_death <- selected_data |>
#   filter(state_abbreviation %in% c("AZ", "KS", "ID", "FL", "GA")) # removed UTAH

#median_value <- median(selected_data$alcohol_impaired_driving_deaths_raw_value, na.rm = TRUE)

# selected_data <- selected_data |>
#   mutate(Counties = ifelse(alcohol_impaired_driving_deaths_raw_value > median_value, 
#                                          "High Alcohol Impaired Deaths", 
#                                          "Low Alcohol Impaired Deaths"))

high_threshold <- 0.3364
low_threshold <- 0.1818

# Create a new variable based on the thresholds
selected_data_first_and_third <- selected_data %>%
  filter(alcohol_impaired_driving_deaths_raw_value > high_threshold | alcohol_impaired_driving_deaths_raw_value < low_threshold) |>
  mutate(Counties = case_when(
    alcohol_impaired_driving_deaths_raw_value > high_threshold ~ "High Alcohol Impaired Deaths Rate",
    alcohol_impaired_driving_deaths_raw_value < low_threshold ~ "Low Alcohol Impaired Deaths Rate",
    TRUE ~ "Medium Alcohol Impaired Deaths"
  ))

ggplot(selected_data_first_and_third, aes(x = median_household_income_raw_value, y = excessive_drinking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  labs(title = 'Median Household Income vs Excessive Drinking',
       x = 'Median Household Income',
       y = 'Excessive Drinking Rate') +
  scale_color_manual(values = c("High Alcohol Impaired Deaths Rate" = "darkred", "Low Alcohol Impaired Deaths Rate" = "darkgreen"))

```


## visualizing adult smoking rate vs drug overdose deaths, where % of people under 18 in poverty is confounding 

```{r}
summary(selected_data$drug_overdose_deaths_raw_value)

high_threshold2 <- 35.75
low_threshold2 <- 17.71

selected_data_first_and_third2 <- selected_data %>%
  filter(drug_overdose_deaths_raw_value > high_threshold2 | drug_overdose_deaths_raw_value < low_threshold2) |>
  mutate(Counties = case_when(
    drug_overdose_deaths_raw_value > high_threshold2 ~ "High Drug Overdose rate",
    drug_overdose_deaths_raw_value < low_threshold2 ~ "Low Drug Overdose rate",
    TRUE ~ "Medium Alcohol Impaired Deaths"
  ))

ggplot(selected_data_first_and_third2, aes(x = children_in_poverty_raw_value, y = adult_smoking_raw_value, color = Counties)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', aes(group = Counties), se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Counties with high overdose rates have high rate of current smokers",
       x = '% of people under 18 in poverty',
       y = '% of adult current smokers') +
  scale_color_manual(values = c("High Drug Overdose rate" = "darkred", "Low Drug Overdose rate" = "darkgreen"))

```

## visualizing adult smoking rate vs drug overdose deaths, where median income is confounding


9 observations that is greater than 100. Examine the numerator and the denominator--

```{r}
# split by quartiles
summary(selected_data$median_household_income_raw_value)
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
   # 4.86   17.71   25.25   28.98   35.75  148.51    1244 

selected_data$`Income level` <- cut(selected_data$median_household_income_raw_value,
                                  breaks = c(-Inf, 52610, 60986, 70897, Inf),
                                  labels = c("Low", "Lower-Middle", "Upper-Middle", "High"),
                                  right = TRUE, na.rm=TRUE)

ggplot(selected_data, aes(x = adult_smoking_raw_value, y = drug_overdose_deaths_raw_value)) +
  geom_point(size = 2, alpha = 0.5, aes(color = `Income level`)) +
  geom_smooth(method = 'lm', se = TRUE) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Counties with high smoking rate have high drug overdose deaths",
       x = 'Adult Smoking Rate',
       y = 'Drug Overdose Deaths per 100,000 population') +
  scale_color_manual(values = c("Low" = "yellow", "Lower-Middle" = "lightgreen", "Upper-Middle" = "yellowgreen", "High" = "darkgreen")) 


```

## visualizing excessive drinking vs median household income

```{r}
# selected_data |>
#   mutate(income_level = )

# at different states, you have different interpretations of what is considered low income, lower-middle
new_df$`Income level` <- cut(new_df$median_household_income_raw_value,
                                  breaks = c(-Inf, 52610, 60986, 70897, Inf),
                                  labels = c("Low", "Lower-Middle", "Upper-Middle", "High"),
                                  right = TRUE, na.rm=TRUE)

new_df <- new_df %>%
  drop_na(excessive_drinking_raw_value, alcohol_impaired_driving_deaths_raw_value)

# go back on this....based on akshay's comments 
# include entries with at least 3000
fifth_percentile <- quantile(new_df$population_raw_value, probs = 0.05)
print(fifth_percentile) # 2672

new_data <-new_df |>
  filter(population_raw_value > 3000)

p <- ggplot(new_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(size = 2, alpha = 0.5, aes(color = `Income level`)) +
  geom_smooth(method = 'lm', se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  scale_x_continuous(labels=scales::percent_format()) +
  labs(title = "Counties with higher median income have higher proportion of adults that excessively drink",
       x = '% of adults that excessively drink',
       y = '% of driving deaths with alcohol involvement') +
  scale_color_manual(values = c("Low" = "lightcyan1", "Lower-Middle" = "lightskyblue2", "Upper-Middle"= "cadetblue", "High" = "darkblue")) 

#install.packages("textGrob")
# library(grid)
# footnote <- textGrob("Restricting to counties with at least a population of 3000 ", 
#                      x = unit(1, "npc") - unit(2, "mm"), 
#                      y = unit(2, "mm"), just = c("right", "bottom"), 
#                      gp = gpar(fontsize = 8, col = "darkgray"))
# 
# p + annotation_custom(footnote)

p + labs(caption = "*Restricting to counties with population > 3000 ")

# counties with higher median income tend to have higher percentage of adults that binge drink leading to higher rate of driving fatalities from alcohol

# lm(alcohol_impaired_driving_deaths_raw_value~excessive_drinking_raw_value, data=selected_data)
# 
# model <- lm(alcohol_impaired_driving_deaths_raw_value ~ excessive_drinking_raw_value, data = selected_data)
# 
# # Extract the coefficients
# coefficients <- coef(model)
# intercept <- coefficients[1]
# slope <- coefficients[2]

```

## exploring pop_den vs num of car crash deaths--is there a linear relationship?

```{r}
ggplot(new_data, aes(x = pop_den, y = alcohol_impaired_driving_deaths_denominator)) +
  scale_x_log10() +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', se = TRUE) +
  labs(x = 'Population Density', y = 'Number of car crashes (log scale)')

ggplot(new_data, aes(x = pop_den, y = alcohol_impaired_driving_deaths_raw_value))  +
  scale_x_log10() +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = 'lm', se = TRUE) +
  labs(x = 'Population density (log scale)', y = 'alcohol impaired death rate')

new_data |> filter(pop_den > 18) |> select( alcohol_impaired_driving_deaths_numerator, alcohol_impaired_driving_deaths_denominator, pop_den)

new_data |> filter(pop_den < 18) |> select( alcohol_impaired_driving_deaths_numerator, alcohol_impaired_driving_deaths_denominator, pop_den)
```

## faceting excessive drinking vs % driving deaths with alcohol invovlement-- graph above by `Population Density`

```{r}
new_df$`Income level` <- cut(new_df$median_household_income_raw_value, breaks = c(-Inf, 52610, 60986, 70897, Inf), labels = c("Low", "Lower-Middle", "Upper-Middle", "High"), right = TRUE, na.rm=TRUE)

# 4 facets
new_df$`Population Density` <- cut(new_df$pop_den, breaks=c(-Inf, 6.145, 17.230, 45.916, Inf), labels = c("Low Population Density", "Moderate-low density", "Moderate-high density", "High Population Density"))

# 2 facets
new_df$`Population Density` <- cut(new_df$pop_den, breaks=c(-Inf, 17.230, Inf), labels = c("Low Population Density", "High Population Density"))

# new_df$`Population Density` <- cut(new_df$pop_den, breaks=c(-Inf, 6.145, 17.230, 45.916, Inf), labels = c("Low Population Density", "Moderate-low density", "Moderate-large density", "Large Population Density"))

new_df <- new_df %>%
  drop_na(excessive_drinking_raw_value, alcohol_impaired_driving_deaths_raw_value)

# go back on this....based on akshay's comments 
# include entries with at least 3000
fifth_percentile <- quantile(new_df$population_raw_value, probs = 0.05)


new_data <-new_df |>
  filter(population_raw_value > 3000)

p <- ggplot(new_data, aes(x = excessive_drinking_raw_value, y = alcohol_impaired_driving_deaths_raw_value)) +
  geom_point(size = 2, alpha = 0.5, aes(color = `Income level`)) +
  geom_smooth(method = 'lm', se = TRUE) +
  scale_y_continuous(labels=scales::percent_format()) +
  scale_x_continuous(labels=scales::percent_format()) +
  facet_wrap(~`Population Density`, scales = "free_y", ncol = 2) +
  labs(title = "*Counties with higher median income have higher proportion of adults that excessively drink",
       x = '% of adults that excessively drink',
       y = '% of driving deaths with alcohol involvement') +
  scale_color_manual(values = c("Low" = "lightcyan1", "Lower-Middle" = "lightskyblue2", "Upper-Middle"= "cadetblue", "High" = "darkblue")) 

#"Upper-Middle" = "cadetblue", "lightskyblue2"

#install.packages("textGrob")
library(grid)
# footnote <- textGrob("Restricting to counties with population > 3000 ", 
#                      x = unit(0.85, "npc") - unit(2, "mm"), 
#                      y = unit(1, "mm"), just = c("right", "bottom"), 
#                      gp = gpar(fontsize = 8, col = "darkgray"))

p + labs(caption = "*Restricting to counties with population > 3000")


```

Only low population density have 0% and 100% alcohol impaired deaths (for the most part). Income-level is not as much as a confounding variable for large population density

## investigating alcohol involvement deaths that are 0 and 100%

```{r}
zero_and_one_alc <-selected_data |>
  filter(alcohol_impaired_driving_deaths_raw_value != c(0, 1))

zero_and_one_alc |>
  select(population_raw_value) |>
  summary()

# population_raw_value for 0 and 1 for alcohol involvement deaths
#  Min.   :  233       
#  1st Qu.: 2174       
#  Median : 4022       
#  Mean   : 5854       
#  3rd Qu.: 7504       
#  Max.   :27936   

# population_raw_value for all data 
#  Min.   :       51   
#  1st Qu.:    11333   
#  Median :    26941   
#  Mean   :   315578   
#  3rd Qu.:    75799   
#  Max.   :333287557   

zero_alcohol_imp_deaths <- selected_data |>
  filter(alcohol_impaired_driving_deaths_raw_value == 0)

summary(zero_alcohol_imp_deaths$population_raw_value)

one_hund_alcohol_imp_deaths <- selected_data |>
  filter(alcohol_impaired_driving_deaths_raw_value == 1) 

summary(one_hund_alcohol_imp_deaths$population_raw_value)

  


```

0 and 1 for generally less pouplated counties

## visualizing alcohol involvement deaths that are 0%, analyzing corresponding numerator and denominator

```{r}
# Look at population, income, location, racial mix
# look at numerator and denominator of alcohol_impaired_driving_deaths

# The numerator is the total number of alcohol-impaired motor vehicle crash deaths in the 5-year period.
# The denominator is the total number of motor vehicle crash deaths in the 5-year period.

examine_zero_alcohol_imp <- selected_data |>
  # select(alcohol_impaired_driving_deaths_numerator, 
  #        alcohol_impaired_driving_deaths_denominator,
  #        alcohol_impaired_driving_deaths_raw_value, population_raw_value) |>
  filter(alcohol_impaired_driving_deaths_raw_value == 0 & population_raw_value > 3000) |>
  select(alcohol_impaired_driving_deaths_denominator, population_raw_value, name, state_abbreviation)

# all numerator is 0 and denminator is < 20 (total number of motor vehicle crash deaths)


```

So I see there is no NA for numerator or denominator of alcohol_impaired. The numerator represents total number of alcohol-impaired motor vehicle crash deaths in the 5-year period. The denominator is total number of motor vehicle crash deaths in the 5-year period The numerator is all 0's and denominators are all very small (<20). I thought a very small population could explain the small number of motor vehicle crash deaths but the population are all decently sized I think (50% are < 7000)? There are several that are from the least populated counties in the state. But I find it hard to believe that there are < 20 crash deaths in the span of 5 years even for these counties. What do you think?

## Modeling

```{r}
gam_mod<-gam(drug_overdose_deaths_raw_value~s(median_household_income_raw_value) + excessive_drinking_raw_value, data=selected_data)

summary(gam_mod)
```