---
title: "Mixed_effects_model"
author: "Jodie"
date: "2024-07-16"
output: html_document
---
## Create a new col state_income (median) for all counties

```{r}
library(pdp)
library(vip)
library(rpart.plot)
library(rpart)
library(caret)
library(tidyverse)
library(janitor)
# just extracting the state median income
income_state <- read.csv("data/analytic_data2024.csv") |> 
  slice(-1) |> 
  clean_names() |> 
  rename(
    Drug_Overdose_Deaths = drug_overdose_deaths_raw_value,
    Alcohol_Impaired_Driving_Deaths = alcohol_impaired_driving_deaths_raw_value,
    Unemployment_Rate = unemployment_raw_value,
    Income_Inequality = income_inequality_raw_value,
    Highschool_Comp_Rate = high_school_completion_raw_value,
    Adult_Smoking_Rate = adult_smoking_raw_value,
    Excessive_Drinking_Rate = excessive_drinking_raw_value,
    Median_House_Income = median_household_income_raw_value,
    College_Education_Rate = some_college_raw_value,
    Children_In_Poverty = children_in_poverty_raw_value,
    Single_Parent_House = children_in_single_parent_households_raw_value,
    Social_Associations = social_associations_raw_value,
    Pop_Under_18 = x_below_18_years_of_age_raw_value,
    Pop_Over_65 = x_65_and_older_raw_value,
    NH_Black = x_non_hispanic_black_raw_value,
    American_Or_Alaska_Native = x_american_indian_or_alaska_native_raw_value,
    Asian_Pop = x_asian_raw_value,
    Pacific_Islander = x_native_hawaiian_or_other_pacific_islander_raw_value,
    Hispanic_Pop = x_hispanic_raw_value,
    NH_White = x_non_hispanic_white_raw_value,
    English_Profic = x_not_proficient_in_english_raw_value,
    Female_Pop = x_female_raw_value,
    Rural_Pop = x_rural_raw_value
  ) |> as_tibble() |>
  filter(county_fips_code=="000" & x5_digit_fips_code != "00000") |>
  select(name, state_abbreviation, Median_House_Income) |>
  print(n=52)
  
# median_household_income_numerator is empty


final_data <- read.csv("data/final_data.csv") 

# append state information to health_data

```

## Making Random Rorest Model for Drug Overdose rate and implementing cross fold validation
## go back on this 

```{r}
library(tidyverse)
library(glmnet)
library(broom)
library(ggfortify)  

health <- read_csv("data/final_data.csv")

# remove population density--15.36785
# with population density
drugs_df <- health |> 
  select(-name, -...1, -population_density, -population, -alcohol_impaired_driving_deaths, -homicides, -excessive_drinking) 

drugs_df <- health |> # select 5 predictors 
  select(unemployment, median_household_income, children_in_single_parent_households, adult_smoking, disconnected_youth, drug_overdose_deaths, homicides) 

drugs_df <- health |> # select 5 predictors 
  select(unemployment, median_household_income, adult_smoking, suicides, drug_overdose_deaths) 



drugs_df |>  summarise(across(everything(), ~ sum(is.na(.))))


drugs_df <- na.omit(drugs_df)

response <- drugs_df$drug_overdose_deaths
predictors <- as.matrix(drugs_df |>  select(-drug_overdose_deaths))

set.seed(123)  # For reproducibility
# cv_fit <- cv.glmnet(predictors, response, alpha = 1, nfolds = 10)
# 
# print(cv_fit)
# 
# plot(cv_fit)

# cv(random forest model)

# finetune mtry (num of predictors at each split)


train<-drugs_df |>
  slice_sample(prop=0.5)
test<-drugs_df|>
  anti_join(train)

x_train<- train |>
  select(-drug_overdose_deaths) |>
  as.matrix()

x_test <- test |>
  select(-drug_overdose_deaths) |>
  as.matrix()

library(caret)
library(tidyverse)
library(ranger)
library(randomForest)

rf_tune_grid <- 
  data.frame(mtry = seq(3, 5, 8), 
           splitrule = "variance",
           min.node.size = 5)
set.seed(123)
drug_overdose_rf <- train(drug_overdose_deaths ~ ., data = train,
        method = "ranger", num.trees = 500,
        trControl = trainControl(method = "cv", number = 10),
        tuneGrid = rf_tune_grid, importance="impurity")

drug_overdose_rf # gives in sample evaluation 12.08305 (average across all 10 test folds)
predictions = predict(drug_overdose_rf, newdata=test) # predicted drug overdose deaths for all the counties


RMSE_test = sqrt(mean((test$drug_overdose_deaths - predictions)^2)) # out of sample evaluation is 13.46791
RMSE_test

# limiting 5 to 6 variables

library(vip)
var_imp<-varImp(drug_overdose_rf)
plot(var_imp)

require(caret)
require(randomForest)
drug_overdose_rf |>
  vip()

varImp.randomForest(drug_overdose_rf)
varImplot(drug_overdose_rf)


# library(ranger)
# drug_overdose_rf2 <- ranger(drug_overdose_deaths ~ ., 
#                             num.trees = 500, importance = "impurity", data = health)
# drug_overdose_rf2
# 
# library(vip)
# vip(drug_overdose_rf)
# drug_overdose_rf$
# varImpPlot(drug_overdose_rf)

```

# partial dependence plots using random forest model

```{r}
# eda 
library(caret)
library(pdp)
library(gridExtra)
# pdp_hispanic<-drug_overdose_rf |>
#   pdp::partial(pred.var="hispanic") |>
#   autoplot()

# final_drug_rf <-ranger(drug_overdose_deaths ~ ., data = train, num.trees=500, mtry=3, min.node.size=5, importance="impurity")
# final_drug_rf |>
#   pdp::partial(pred.var="hispanic") |>
#   autoplot()

# pdp_american<-drug_overdose_rf |>
#   pdp::partial(pred.var="american_indian_or_alaska_native") |>
#   autoplot()

# cutoff 8%
pdp_unemployment<-drug_overdose_rf |>
  pdp::partial(pred.var="unemployment") |>
  autoplot()

pdp_unemployment<-pdp_unemployment +
  labs(x = "unemployment rate", y="predicted drug overdose deaths")

pdp_adult_smoking<-drug_overdose_rf |>
  pdp::partial(pred.var="adult_smoking") |>
  autoplot()

pdp_suicides<-drug_overdose_rf |>
  pdp::partial(pred.var="suicides") |>
  autoplot()

pdp_income<-drug_overdose_rf |>
  pdp::partial(pred.var="median_household_income") |>
  autoplot()

#plot_hispanic <- autoplot(pdp_hispanic)
#plot_hispanic <- autoplot(pdp_hispanic)
plot_suicides <- autoplot(pdp_suicides)
plot_income <- autoplot(pdp_income)
plot_unemployment <- autoplot(pdp_unemployment)
plot_adult_smoking <- autoplot(pdp_adult_smoking)



grid_pdp<-grid.arrange(plot_unemployment, plot_adult_smoking, plot_income, ncol = 3)
grid_pdp

# plot_adult_smoking + labs(x = 'Adsd') to relabel the axes

```

```{r}
drugs_df2 <- drugs_df |>
  mutate(employment = 1-unemployment)
train<-drugs_df2 |>
  slice_sample(prop=0.5)
test<-drugs_df2|>
  anti_join(train)

x_train<- train |>
  select(-drug_overdose_deaths) |>
  as.matrix()

x_test <- test |>
  select(-drug_overdose_deaths) |>
  as.matrix()

library(caret)
library(tidyverse)
library(ranger)
library(randomForest)

rf_tune_grid <- 
  data.frame(mtry = seq(3, 5, 8), 
           splitrule = "variance",
           min.node.size = 5)
set.seed(123)
drug_overdose_rf <- train(drug_overdose_deaths ~ ., data = train,
        method = "ranger", num.trees = 500,
        trControl = trainControl(method = "cv", number = 10),
        tuneGrid = rf_tune_grid, importance="impurity")

drug_overdose_rf # gives in sample evaluation 12.08305 (average across all 10 test folds)
predictions = predict(drug_overdose_rf, newdata=test) # predicted drug overdose deaths for all the counties


RMSE_test = sqrt(mean((test$drug_overdose_deaths - predictions)^2)) # out of sample evaluation is 13.46791
RMSE_test

pdp_american<-drug_overdose_rf |>
  pdp::partial(pred.var="american_indian_or_alaska_native") |>
  autoplot()

# cutoff 8%
pdp_unemployment<-drug_overdose_rf |>
  pdp::partial(pred.var="employment") |>
  autoplot()

pdp_adult_smoking<-drug_overdose_rf |>
  pdp::partial(pred.var="adult_smoking") |>
  autoplot()

pdp_children<-drug_overdose_rf |>
  pdp::partial(pred.var="children_in_single_parent_households") |>
  autoplot()

plot_hispanic <- autoplot(pdp_hispanic)
plot_american_indian_or_alaska_native <- autoplot(pdp_american)
plot_unemployment <- autoplot(pdp_unemployment)
plot_adult_smoking <- autoplot(pdp_adult_smoking)



grid_pdp<-grid.arrange(plot_american_indian_or_alaska_native,
             plot_unemployment, plot_adult_smoking, pdp_children, ncol = 2)
grid_pdp

```

## Mixed effects Model for Drug Overdose where random intercept is urban/rural counties (clustering by urban and rural counties)

```{r setup, include=FALSE}

# mixed effects model
# drop homicides for later

# create a new col called income_level with 4 levels--low, high, middle, upper middle, lower middle
# effect of income level changes by state --random slope

health_data <- read.csv("data/analytic_data2024.csv") |> 
  slice(-1) |> 
  clean_names() 

sel_data<-health_data |>
  rename(
    Drug_Overdose_Deaths = drug_overdose_deaths_raw_value,
    Alcohol_Impaired_Driving_Deaths = alcohol_impaired_driving_deaths_raw_value,
    Unemployment_Rate = unemployment_raw_value,
    Income_Inequality = income_inequality_raw_value,
    Highschool_Comp_Rate = high_school_completion_raw_value,
    Adult_Smoking_Rate = adult_smoking_raw_value,
    Excessive_Drinking_Rate = excessive_drinking_raw_value,
    Median_House_Income = median_household_income_raw_value,
    College_Education_Rate = some_college_raw_value,
    Children_In_Poverty = children_in_poverty_raw_value,
    Single_Parent_House = children_in_single_parent_households_raw_value,
    Social_Associations = social_associations_raw_value,
    Pop_Under_18 = x_below_18_years_of_age_raw_value,
    Pop_Over_65 = x_65_and_older_raw_value,
    NH_Black = x_non_hispanic_black_raw_value,
    American_Or_Alaska_Native = x_american_indian_or_alaska_native_raw_value,
    Asian_Pop = x_asian_raw_value,
    Pacific_Islander = x_native_hawaiian_or_other_pacific_islander_raw_value,
    Hispanic_Pop = x_hispanic_raw_value,
    NH_White = x_non_hispanic_white_raw_value,
    English_Profic = x_not_proficient_in_english_raw_value,
    Female_Pop = x_female_raw_value,
    Rural_Pop = x_rural_raw_value
  ) 

selected_data <- sel_data|> 
  select(
    name, state_abbreviation, x5_digit_fips_code, x5_digit_fips_code, Drug_Overdose_Deaths, Alcohol_Impaired_Driving_Deaths,
    Unemployment_Rate, Income_Inequality, Highschool_Comp_Rate, 
    Adult_Smoking_Rate, Excessive_Drinking_Rate, Median_House_Income,
    College_Education_Rate, Children_In_Poverty, Single_Parent_House,
    Social_Associations, Pop_Under_18, 
    Pop_Over_65, NH_Black, 
    American_Or_Alaska_Native, Hispanic_Pop, 
    Female_Pop, Rural_Pop, population_raw_value
  ) |>
  mutate(x5_digit_fips_code=ifelse(nchar(x5_digit_fips_code) == 4, paste0("0", x5_digit_fips_code), x5_digit_fips_code)) |>
  mutate(across(c(Drug_Overdose_Deaths, Alcohol_Impaired_Driving_Deaths,
    Unemployment_Rate, Income_Inequality, Highschool_Comp_Rate, 
    Adult_Smoking_Rate, Excessive_Drinking_Rate, Median_House_Income,
    College_Education_Rate, Children_In_Poverty, Single_Parent_House,
    Social_Associations, Pop_Under_18, 
    Pop_Over_65, NH_Black, 
    American_Or_Alaska_Native, Hispanic_Pop, 
    Female_Pop, Rural_Pop, population_raw_value), 
    as.numeric
  ))

  
  
# selected_data <- selected_data |> 
#   mutate(across(
#     everything(),
#     as.numeric
#   ))

library(tigris)
#us_counties df is strictly 5 numbers while selected_data has 4 digits for nums less than 10000

us_counties <- counties(cb = TRUE) # US COUNTIES, census_df have CA but selected_data doesn't have it (06037)
census_df <- us_counties |> # issue with census df not having ca counties
  select(x5_digit_fips_code=GEOID, land_area=ALAND)

# Inner join: only rows with matching x5_digit_fips_code will be included
new_sel <- merge(selected_data, census_df, by = "x5_digit_fips_code") |>
  mutate(land_area_km = land_area / 1000000) |>
  mutate(pop_den = population_raw_value / land_area_km) |>
  mutate(urban_pop=1-Rural_Pop)
  

quantiles <- quantile(new_sel$urban_pop, probs = c(0, 0.33, 0.66, 1))

# Display the quantiles
print(quantiles)

new_df <- new_sel |>
  mutate(urban_level = ifelse(urban_pop==0, "rural", ifelse(urban_pop<0.52, "suburban", "urban"))) 

# high school--completed at least high school diploma
# some college--completed high school and pursued education after (not necessarily have to complete it)
# union--% that did not complete (1-high school)

# % of people under 18 in poverty (synonymous with income inequality)
new_drugsdf <- new_df |>
  select(-Alcohol_Impaired_Driving_Deaths, -Excessive_Drinking_Rate, 
         -geometry, -land_area_km, -population_raw_value, -pop_den, 
         -Rural_Pop, -urban_pop, -land_area, -Highschool_Comp_Rate,
         -Pop_Under_18, -Pop_Over_65, -Children_In_Poverty)|>
  mutate(across(c(Unemployment_Rate, Income_Inequality, Adult_Smoking_Rate, 
                  Median_House_Income, College_Education_Rate, 
                  Social_Associations, NH_Black, American_Or_Alaska_Native, Hispanic_Pop, Female_Pop), scale)) |>
  mutate(income_ineq_level=ifelse(Income_Inequality<0.123, "low", "high")) |>
  mutate(urban_num=as.numeric(factor(urban_level, levels = c("rural", "suburban", "urban")))) |>
  drop_na(Drug_Overdose_Deaths) # omitted 20 rows?


library(lme4)
random_urban <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level), data=new_drugsdf)

fixed_no_urban <-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop, data=new_drugsdf)

random_state <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|state_abbreviation), data=new_drugsdf)

ggplot(aes(x = type, y = rmse)) + 
  geom_point(size = 4) +
  stat_summary(fun = mean, geom = "point", 
               color = "red", size = 4) + 
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               color = "red", width = 0.2)








# urban_num has no effect since CI includes 0
fixed_urban<-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + urban_num, data=new_drugsdf)

confint(fixed_urban)

random_state <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|state_abbreviation), data=new_drugsdf)

library(phia)
model_inter<-aov(Drug_Overdose_Deaths~factor(income_ineq_level)*factor(urban_level), data=new_drugsdf)
inter_plot<-interactionMeans(model_inter)
plot(inter_plot)
par(mfrow=c(1, 2)) # no interaction between income inequality level and urban_level; 


anova(random_urban, fixed_no_urban) # likelihood ratio test--> look for lower AIC and lower BIC 

# random_urban_one <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + (1 |state_abbreviation), data=new_drugsdf)

random_state <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|state_abbreviation), data=new_drugsdf)

# intercept and slope of lines should be different 
# maybe median_house_income * urban_yes

summary(random_urban)

# use new_df 
#(1 + income | state) # and income could be continuous, making state a random variable

#(1|state)


new_csv<-my_csv |> 
  left_join(select(income_state, -name), by = "state_abbreviation") |>
  rename(state_income=Median_House_Income) |>
  mutate(relative_income = median_household_income/state_income) |>
  mutate(categ_rel = c)

# use percentiles of relative_income 
# aggregate by state, and create histogram? 

# View()

# wouldn't worry about this

final_data<-read_csv("data/final_data.csv")

# (1 | state)
# random intercept for state allows baseline drug overdose deaths to vary by state, to account for state level
# variability
# or do we want 

  
```

## Compare regression models using AIC

```{r cars}
# drugs_df2 <- new_drugsdf |>
#   mutate(employment = 1-unemployment)
new_drugsdf <- new_drugsdf |>
  select(Drug_Overdose_Deaths, Unemployment_Rate, Income_Inequality, Median_House_Income, Adult_Smoking_Rate, College_Education_Rate, Social_Associations, Female_Pop, urban_level, urban_num, state_abbreviation) |>
  drop_na()

train<-new_drugsdf |>
  slice_sample(prop=0.5)
test<-new_drugsdf|>
  anti_join(train)

x_train<- train |>
  select(-Drug_Overdose_Deaths) |>
  as.matrix()

x_test <- test |>
  select(-Drug_Overdose_Deaths) |>
  as.matrix()

library(tidyverse)
library(ranger)
library(randomForest)

library(caret)

fixed_no_urban <-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop, data=new_drugsdf)

fixed_urban <-lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + urban_num, data=new_drugsdf)

random_urban <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level), data=new_drugsdf)

random_state <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + urban_num + (1|state_abbreviation), data=new_drugsdf)

# random_state with log transformation
random_state_log <-lmer(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + urban_num + (1|state_abbreviation), data=new_drugsdf)

random_urban_log <-lmer(log(Drug_Overdose_Deaths)~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level), data=new_drugsdf)

# select random_state

arm::plot(fitted(random_state), residuals(random_state))
# random_state_inter_income_smoking <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income*Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + urban_num + (1|state_abbreviation), data=new_drugsdf)

library(car)
vif(random_state)
# strong correlation between median household income and adult smoking rate 

# state is random intercept, interact female with urban_num
# inter_fem_urban_num <-lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop*urban_num + (1|state_abbreviation), data=new_drugsdf)

AIC(fixed_no_urban, fixed_urban, random_urban, random_state) # random_state does the best
confint(random_state, method="boot")
ranef_model <- ranef(random_state, condVar=TRUE) # random_effects--state intercept

library(arm)
se_ranef <- se.ranef(random_state)

# Extract standard deviations for the random effects
sd_ranef <- lapply(se_ranef, function(x) sqrt(diag(x)))

# Calculate confidence intervals for random effects
conf_intervals <- lapply(ranef_model, function(re, sd_re) {
  re_lwr <- re - 1.96 * sd_re
  re_upr <- re + 1.96 * sd_re
  return(data.frame(Lower=as.vector(re_lwr), Upper=as.vector(re_upr)))
}, sd_ranef)


# grpah wtih the summary estimates and coefficients
# AIC score has to be at least 2 units lower compared to the other model for it to be significant enough.

# visualization of the AIC (for later) --delta aic 

# install.packages("lmerTest")
library(lmerTest)

model_random_state = summary(random_state)
coefficients <- model_random_state$coefficients[2:10, 1] 
conf_intervals <- confint(random_state, level = 0.95, oldNames=FALSE)

coef_df <- data.frame(
  Predictor = c('Unemployment Rate', 'Income Inequality',
                'Median Income', 'Smoking Rate', 'College Education Rate', 
                'Social Associations', '% Female', 'Urbanization'),
  Estimate = coefficients,
  #StdError = coefficients[, "Std. Error"],
  ConfLow = conf_intervals[4:11, "2.5 %"],
  ConfHigh = conf_intervals[4:11, "97.5 %"]
)
#coef_df<-coef_df[c(-2, -3, -4), ]
new_row <- data.frame(
  Predictor='Unemployment Rate',
  Estimate = 1.8850,
  ConfLow=1.0461464,
  ConfHigh=2.7238723
)
coef_df<-rbind(coef_df, new_row) 

coef_df <- coef_df[2:10, ]

ggplot(coef_df, aes(x = Estimate, y = Predictor, xmin = ConfLow, xmax = ConfHigh)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(height = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Fixed Effects Estimates for Drug Overdose Deaths",
       x = "Coefficient",
       y = "Predictors") +
  labs(y = NULL) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),                   # X-axis title text size
    axis.title.y = element_text(size = 14),                   # Y-axis title text size
    axis.text.x = element_text(size = 12),                    # X-axis text size
    axis.text.y = element_text(size = 12),                    # Y-axis text size
    legend.text = element_text(size = 12)                      # Legend text size
  )
  

abline(v = 0, col = "black", lwd = 2, lty = 2)

install.packages("lme4")
# install.packages("broom.mixed")
library(lme4)
library(broom.mixed)
example(lmer)

# create get estimates effects of state 

# exclude fixed effects
# create a plot that visualizes the different intercepts for random state and CI

# conf.low and conf.high gives 95% confidence interval
tidy_model <- tidy(random_state, effects="ran_vals", conf.int = TRUE)
print(tidy_model)

ggplot(tidy_model, aes(x = estimate, y = level, ymin = conf.low, ymax = conf.high)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(height = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Random effects for Drug Overdose Deaths",
       x = "Coefficient",
       y = "Predictors") +
  labs(y = NULL) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),                   # X-axis title text size
    axis.title.y = element_text(size = 14),                   # Y-axis title text size
    axis.text.x = element_text(size = 12),                    # X-axis text size
    axis.text.y = element_text(size = 12),                    # Y-axis text size
    legend.text = element_text(size = 12)                      # Legend text size
  )

library(forcats)
tidy_model$level <- fct_reorder(tidy_model$level, tidy_model$estimate)

ggplot(tidy_model, aes(y = estimate, x = level, ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbar(width = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Random Effects Estimates for Drug Overdose Deaths",
       y = "Intercept Estimates",
       x = "State") +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),    # X-axis title text size
    axis.title.y = element_text(size = 14),    # Y-axis title text size
    axis.text.x = element_text(size = 6.7),     # X-axis text size
    axis.text.y = element_text(size = 12)      # Y-axis text size
  )



# make a 

# estimates for variance components of random effects--can look at the interpretation
tidy_ran_pars <- tidy(random_state, effects = "ran_pars")
print(tidy_ran_pars)

# another way of making visualization
library(lme4)
example(lmer)
rr <- ranef(random_state)
class(rr) ## ranef.mer
methods(class = "ranef.mer")

head(as.data.frame(rr), 3)
lattice::dotplot(rr)

summary(random_state)




summary(inter_social_urban_num)
confint(inter_social_urban_num)
confint(random_state)

## Evaluate the AIC for regression models


################training#####################
# Define the training control for cross-validation

new_drugsdf <- new_drugsdf |>
  mutate(test_fold = sample(rep(1:5, length.out = n())))

rf_tune_grid <- 
  data.frame(mtry = seq(3, 5, 8), 
           splitrule = "variance",
           min.node.size = 5)
set.seed(123)
drug_overdose_rf <- train(Drug_Overdose_Deaths ~ ., data = train,
        method = "ranger", num.trees = 500,
        trControl = trainControl(method = "cv", number = 5),
        tuneGrid = rf_tune_grid, importance="impurity")

drug_overdose_rf # gives in sample evaluation 12.08305 (average across all 10 test folds)
predictions = predict(drug_overdose_rf, newdata=test)
test_rf = sqrt(mean((test$Drug_Overdose_Deaths - predictions)^2)) # out of sample evaluation is 13.46791
test_rf

# get_test_pred <- function(k) {
#   test_data <- new_drugsdf |>  filter(test_fold == k)                     # get test and training data
#   train_data <- new_drugsdf |>  filter(test_fold != k)
#   test_x <- as.matrix(select(test_data, -Drug_Overdose_Deaths))            # get test and training matrices
#   train_x <- as.matrix(select(train_data, -Drug_Overdose_Deaths))
#   
#   lm_fixed <- lm(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop, data = train_data)                  # fit models to training data
#   lm_rand_urban <- lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level), data=train_data)
#   lm_rand_state <- lmer(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|state_abbreviation), data=train_data)
# 
#   # rf_fit <- randomForest(drug_overdose_deaths ~ ., data = train_data, importance = TRUE)
#   tibble(lm_pred = predict(lm_fixed, newdata = test_data),              # return test results
#          lm_rand_urban = predict(lm_rand_urban, newdata = test_data),
#          lm_rand_state = predict(lm_rand_state, newdata = test_data),
#          rf = predict(drug_overdose_rf, newdata=test),
#          rf_pred = predict(rf_fit, newdata = test_data),
#          test_actual = test_data$drug_overdose_deaths,
#          test_fold = k)
# }

train_control <- trainControl(method = "cv", number = 5)


fixed_ml <- train(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop,
                  data=train, method="lm", trControl=train_control)

pred_fixed <- predict(fixed_ml, newdata=test)
test_fixed = sqrt(mean((test$Drug_Overdose_Deaths - pred_fixed)^2)) # out of sample evaluation is 13.46791
test_fixed

loocv_linear <- function(i, formula) {
  
  train_data <- new_drugsdf[-i, ]
  test_data <- new_drugsdf[i, , drop = FALSE]
  
  model <- lmer(formula, data = train_data)
  pred <- predict(model, newdata = test_data)
  
  return((test_data$Drug_Overdose_Deaths - pred)^2)
}

# for bootstrapping
library(boot)
rmse_function <- function(data, indices) {
  boot_data <- data[indices, ] # Sample with replacement
  squared_errors <- sapply(1:nrow(boot_data), loocv_linear)
  return(sqrt(mean(squared_errors)))
}
  

randurb_form<-Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level)

sq_error_randurb<-sapply(1:nrow(new_drugsdf), loocv_mixed, formula=randurb_form)
rmse_randurb=sqrt(mean(sq_error_randurb)) # 14.91057

# Bootstrapping
set.seed(123)
results <- boot(data = new_drugsdf, statistic = rmse_function, R = 1000)

#loocv_mixed(new_drugsdf, randurb_form)


# doesn't work for cross-fold validation so try leave one out cross fold validation
random_urban <- train(Drug_Overdose_Deaths~Unemployment_Rate + Income_Inequality + Median_House_Income + Adult_Smoking_Rate + College_Education_Rate + Social_Associations + Female_Pop + (1|urban_level),
                  data=train, method="lmer", trControl=train_control)

pred_urban_rand <- predict(fixed_ml, newdata=test)
test_urban_rand = sqrt(mean((test$Drug_Overdose_Deaths - pred_urban_rand)^2)) # out of sample evaluation is 13.46791
test_urban_rand # 14.91057 for urban as a random variable



rf_tune_grid <- 
  data.frame(mtry = seq(3, 5, 8), 
           splitrule = "variance",
           min.node.size = 5)
set.seed(123)
drug_overdose_rf <- train(Drug_Overdose_Deaths ~ ., data = train,
        method = "ranger", num.trees = 500,
        trControl = trainControl(method = "cv", number = 10),
        tuneGrid = rf_tune_grid, importance="impurity")

drug_overdose_rf # gives in sample evaluation 12.08305 (average across all 10 test folds)
predictions = predict(drug_overdose_rf, newdata=test) # predicted drug overdose deaths for all the counties


RMSE_test = sqrt(mean((test$drug_overdose_deaths - predictions)^2)) # out of sample evaluation is 13.46791
RMSE_test



```

## New linear graph

You can also embed plots, for example:

```{r pressure, echo=FALSE}
library(lmerTest)

model_random_state = summary(random_state)
coefficients <- c(9.4, 12.473, model_random_state$coefficients[, 1]) 
conf_intervals <- confint(random_state, level = 0.95, oldNames=FALSE)

coef_df <- data.frame(
  Predictor = c('Std Dev estimate', 'Sigma', 'Fixed Intercept', 'Unemployment Rate', 'Income Inequality',
                'Median Income', 'Smoking Rate', 'College Education Rate', 
                'Social Associations', '% Female', 'Urbanization'),
  Estimate = coefficients,
  #StdError = coefficients[, "Std. Error"],
  ConfLow = conf_intervals[, "2.5 %"],
  ConfHigh = conf_intervals[, "97.5 %"]
)
coef_df<-coef_df[c(-2, -3, -4), ]
new_row <- data.frame(
  Predictor='Unemployment Rate',
  Estimate = 1.8850,
  ConfLow=1.0461464,
  ConfHigh=2.7238723
)
coef_df<-rbind(coef_df, new_row)

ggplot(coef_df, aes(x = Estimate, y = Predictor, xmin = ConfLow, xmax = ConfHigh)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_errorbarh(height = 0.2, linewidth = 1.4, color = "red", show.legend = FALSE) +
  geom_point(size = 1.5, shape = 21, fill = "black", show.legend = FALSE) +
  labs(title = "Random Intercept (state) Linear Model for Drug Overdose Deaths",
       x = "Coefficient Estimate",
       y = "Predictors") +
  labs(y = NULL) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 16),      # Title text size
    axis.title.x = element_text(size = 14),                   # X-axis title text size
    axis.title.y = element_text(size = 14),                   # Y-axis title text size
    axis.text.x = element_text(size = 12),                    # X-axis text size
    axis.text.y = element_text(size = 12),                    # Y-axis text size
    legend.text = element_text(size = 12)                      # Legend text size
  )
  
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
