---
title: "Drug overdose and alcohol related deaths"
author: 
- Dennis Campoverde-Lema
- Samantha Parras 
- Jodie Chen 
date: "July 26, 2024"
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Introduction

Are there demographic and social factors that are predictors of drug overdose and alcohol-related incidents (e.g., driving accidents)?

## Data

  The data is collected by the University of Wisconsin Population Health Institute. It assigns a ranking based on health outcomes and health factors to each county in every state. There was originally 770 variables in the data set.

  Since the data set is so big, we decided that we should create another data set with only the variables we want to look at with relation to drug overdose deaths and alcohol impaired driving deaths. 

```{r}
library(janitor)
library(ggplot2)
library(tidyverse)
library(sf)

health_data <- read.csv("data/analytic_data2024.csv") %>%  
  clean_names()

selected_data <- health_data %>% 
  select(
    state_fips_code, county_fips_code, state_abbreviation, name,
    drug_overdose_deaths_raw_value, alcohol_impaired_driving_deaths_raw_value,
    unemployment_raw_value, income_inequality_raw_value, high_school_completion_raw_value, 
    adult_smoking_raw_value, excessive_drinking_raw_value, median_household_income_raw_value,
    some_college_raw_value, children_in_poverty_raw_value, children_in_single_parent_households_raw_value,
    social_associations_raw_value, x_below_18_years_of_age_raw_value, 
    x_65_and_older_raw_value, x_non_hispanic_black_raw_value, 
    x_american_indian_or_alaska_native_raw_value, x_asian_raw_value, 
    x_native_hawaiian_or_other_pacific_islander_raw_value, x_hispanic_raw_value, 
    x_non_hispanic_white_raw_value, x_not_proficient_in_english_raw_value, 
    x_female_raw_value, x_rural_raw_value
  )
```
 
Upon inspection, we found that the variable of drug overdose deaths had lots of N/A values and we want to make sure this is clean as this is one of the main varibales we want to look at. 

```{r}
selected_data %>% 
  mutate(across(everything(), ~ifelse(is.na(.) | . == 0, mean(., na.rm = TRUE), .))) 
```


## Methods
First, we needed to find our variables of interest. We decided to find the p-values 
for drug overdose deaths and alcohol impaired deaths. 


```{r} echo: false}

```

You can add options to executable code like this

```{r}
#| echo: false

```

The `echo: false` option disables the printing of code (only output is displayed).
